Variance reduction for MCMC methods via martingale representations

### 2020-03-11 aap/Revision/vr-mcmc-revision-new.tex
update: variance of vanilla estimator

\def\PE{\mathsf{E}}
\def\PVar{\mathsf{Var}}
\newcommandx{\CPE}[3][1=]{{\mathsf E}_{#1}\left[\left. #2 \, \right| #3 \right]}
\newcommand{\ps}[2]{\langle #1, #2 \rangle}
\def\nset{\mathbb{N}}
\def\rset{\mathbb{R}}
\usepackage[textsize=footnotesize]{todonotes} % pour voir les commentaires

@@

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
\label{eq:qpl}
Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
\label{eq:29032018a5}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
for any measurable function $g$  and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\label{eq:expression-variance}
\PVar_x[\pi_{K,n}^N(f)]=\sum_{k=K+1}^{\infty}\frac{1}{n^2}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
\label{eq:definition-A-s,k}
A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Estimation of coefficients}
The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
We present a generic regression algorithm. Algorithm starts with estimating the functions \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) for $r=N+1,\ldots,N+n.$ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
(the so-called ``training paths''):
\TrainSet=\Bigl\{(X^{(s)}_{0},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
\begin{equation}\label{eq:06042018a1}
for \(r=1,\ldots, n-1,\)
where \(\Psi\) is a  class of  functions on \(\mathbb{R}^d.\)
Next  we estimate the coefficients \( \bar a_{l,k}\) using
the formula
\label{eq:a_est_int}
\widehat a_{l,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat Q_{l}\left(\Phi(x,\xi)\right)\right | \TrainSet\right] \,,
where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
$\Phi_l$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\). Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories. In many cases integration can be done in closed analytical form. For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi(x,\xi)=x-\gamma\mu(x)+\sqrt{\gamma}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
\widehat M_{K,n}^N(f) =\frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
Obviously, \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
does not introduce any bias. By the Jensen inequality and orthonormality of \((\phi_k)_{k\geq 0},\)
\begin{multline*}
\mathsf{E}_x\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \right]
\leq \sum_{k=1}^{K}\frac{1}{n^2}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
\end{multline*}

\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. We use the notations of example \ref{exam:langevin-algorithm}. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

@@

\paragraph{Variance:} An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
under {\bf (H1)} and {\bf (H2)} is derived in Section ??. \denis{Insert} In particular, for any bounded function $f,$
\begin{equation}
\label{eq:var-mc}
\PVar[\pi_n^N(f)]\lesssim (n\gamma)^{-1} \eqsp,
\end{equation}
where $\lesssim$ stands for inequality up to a constant depending on constants $L,$ $m,$  $M$ and $\|f\|_\infty.$
One of the main results of this paper is the following  upper bound for the variance of $\pi_{K,n}^{N}(f).$

@@

Let us sketch the main steps of the proof. First using integration by parts, we prove that

@@

A_{q,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}_x\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_q) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right],

@@

where $F_q(x,Z_1,\ldots,Z_q)=\sum_{r=0}^q f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part} we derive

@@

\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}

@@

\PVar \left( \sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)

@@

\right)\leq C_{K}\gamma^{-1}(1+V(x)), \quad V(x) = 1 + \|x\|^2

@@

for some constant $C_K$ not depending on $q$ and $\gamma$.

@@

Aforementioned lemmas allow us to bound the exponential moment of the additive functional of ULA. Namely, the following theorem holds

\begin{thm}
\label{thm:ula_const_size}
Suppose that ULA kernel \ref{eq:kernel} satisfies assumptions {\bf (H1)} and {\bf (H2)}, and let $X_1,\ldots,X_n,\ldots$ be generated by ULA with constant step size $\gamma$. Let $X_0 = x$ be fixed. Then for any bounded function $g(x) : \mathbb{R}^d \rightarrow \mathbb{R}, |g(x)| \leq M$, it holds
\begin{eqnarray*}
\mathsf{E}\left[\exp\left(-\rho\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] \leq C_1\exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
\end{eqnarray*}
for some absolute constant $C_1$ not depending on $l,p$.
\end{thm}

\begin{proof} Conditions {\bf (H1)} and {\bf (H2)} imply that the Markov kernel $R_\gamma$ is $V-$ergodic (lemma \ref{lem:v_ergodicity}) and satisfies drift condition \ref{lem:drift}. Due to \cite[Theorem~3, Fact 3]{moulines:bound_dif:2019}, we obtain the following bound
\begin{align*}
\mathsf{E}_x\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)\right] \leq \exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
\end{align*}
\sergey{I will put precise constant later}
with constant $\kappa = ...$. Note that
\begin{align*}
\mathsf{E}\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] =
\end{align*}

\begin{align*}
=\mathsf{E}_x\left[\exp{\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)}\right]
\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)}
\end{align*}
Using lemma~\ref{lem:v_ergodicity},
\begin{align*}
|\mathsf{E} g(X_k) - \pi_\gamma(g)| \leq Md_V(\pi_\gamma, \delta_xQ^{1,n}_{\boldsymbol{\gamma}}) \leq CM\rho^k\left(V(x) + \pi_\gamma(V)\right)
\end{align*}
Hence,
\begin{align*}
\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)} \leq \exp{\left(s\gamma CM(V(x) + \pi_\gamma(V))\frac{\rho^l - \rho^{p+1}}{1-\rho}\right)}
\end{align*}
and the statement follows.
\end{proof}

Let us formulate and prove prove analogue of the previous theorem for inhomogeneous Markov chain with step size $\gamma_1,\ldots,\gamma_n,\ldots$. Note that under assumptions $\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty$ and {\bf (H1)} the stationary distribution of ULA-based chain will be equal to $\pi$. We will need additional assumption
\begin{description}
\item[{\bf (H3)}] There exist such constants $\rho > 0, K_3 > 0$ that
\begin{eqnarray*}
d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi) \leq K_3 V(x) \rho^n
\end{eqnarray*}
\end{description}

\begin{thm}
\label{th:exp_mom_ula}
Suppose that
\begin{eqnarray*}
\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty.
\end{eqnarray*}
Under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)}, for any bounded function $g$ on $\mathbb{R}^d$
we have
%\begin{eqnarray*}
%\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
%\end{eqnarray*}
 %and
%\begin{eqnarray*}
%\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
%\end{eqnarray*}
%Moreover,
\begin{eqnarray}
\label{eq:exp_decrease_step}
\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
\end{eqnarray}
for all natural $0<l\leq p<\infty$ with constant $C$ not depending on $l,p.$. Moreover,
\begin{eqnarray}
\label{eq:var_decrease_step}
\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k)\right)} \leq \frac{C_1}{\Gamma_{l+1,l+n}}
\end{eqnarray}
\end{thm}
\begin{proof}

 Note that under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)} it holds due to \cite[Theorem~4]{moulines:bound_dif:2019} that
 \begin{align}
 \label{eq:exp_bounded_dif}
\mathsf{E}_x\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k}) - \mathsf{E}_x \left(\sum\limits_{k=l}^{p}\gamma_{k+1}g(X_k)\right)\right)\right)\right] \leq \exp{\left(s^2 \kappa \sum\limits_{k=l}^{p}\gamma_{k+1}^2\right)} \leq \exp{\left(s^2 \kappa \Gamma^*\right)}
\end{align}
\sergey{This theorem is not yet written in the second paper}
with $\Gamma^* = \sum\limits_{k=1}^{\infty}\gamma_k^2 < \infty$ and $\kappa = ...$. Using the same transformation as in theorem~\ref{thm:ula_const_size}, and using the bound
\begin{align*}
|\mathsf{E}_x f(X_k) - \pi(f)| \leq K_3 M \rho^{\Gamma_{1,k}}V(x)
\end{align*}
we obtain
\begin{align*}
\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right] \leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \sum\limits_{k=l}^{p}\rho^{\Gamma_{1,k}}\gamma_{k+1}\right)} \leq
\end{align*}

>>>>

\def\P{\mathsf{P}}
\def\PE{\mathsf{E}}
\def\PVar{\mathsf{Var}}
\def\PCov{\mathsf{cov}}
\newcommandx{\CPE}[3][1=]{{\mathsf E}_{#1}\left[\left. #2 \, \right| #3 \right]}
\newcommand{\ps}[2]{\langle #1, #2 \rangle}
\def\nset{\mathbb{N}}
\def\rset{\mathbb{R}}
\usepackage[textsize=footnotesize]{todonotes} % pour voir les commentaires

@@

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
\label{eq:29032018a5}
\end{eqnarray}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
\[
\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
\]
for any measurable function $g$  and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar_x[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\frac{1}{n^2}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
\end{equation}
where 
\begin{eqnarray}
\label{eq:definition-A-s,k}
A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N}
\end{eqnarray}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. We use the notations of example \ref{exam:langevin-algorithm}. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

@@

\paragraph{Variance:} In this section we derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain.

\begin{lem} 
\label{lem:variance}
Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$, satisfying {\bf (H1)} and {\bf (H2)} and assume also $n\gamma > C_0$ for some fixed $C_0$. Then it holds
\begin{equation}
\label{eq:var-mc}
\PVar[\pi_n^N(f)]\lesssim \frac{V(x)}{n\gamma} \eqsp,
\end{equation}
with $V(x) = 1+ \|x\|^2$ where $\lesssim$ stands for inequality up to a constant depending on constants $L,$ $m,$  $M$ and $\mathrm{osc}(f)$.
\end{lem}
\begin{proof}
Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic with $V(x) = 1+\|x\|^2$ and $\rho = \rme^{-\kappa\gamma}$ with $\kappa$ specified in lemma~\ref{lem:v_ergodicity}. Due to lemma~\ref{lem:covariance}, 
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})}
\]
and the result of lemma follows from Taylor expansion of denominator.
\end{proof}

One of the main results of this paper is the following  upper bound for the variance of variance-reduced estimator $\pi_{K,n}^{N}(f).$

@@

Let us sketch the main steps of the proof. First using integration by parts we prove that

@@

A_{s,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_s) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right]

@@

where $F_s(x,Z_1,\ldots,Z_s)=\sum_{r=0}^s f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive

@@

\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{s,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}

@@

\PVar_x \left( \sum_{p=1}^{s}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)

@@

@@

for some constant $C$ not depending on $q$ and $V(x) = 1 + \|x\|^2$.

@@

\section{Covariance estimation for $V-$geometrically ergodic Markov chains}\label{sec:appendix_moments}
In this section we assume that $(X_p)_{p \in \nset_0}$ is a $V-$geometrically ergodic Markov chain and prove bounds on the variance of ergodic average $\pi_N^n(f)$ of the form \eqref{eq:29032018a2}. We use the same technique as in \cite{belomestny2019esvm} to control autocovariances for a given Markov chain. We start from auxiliary lemma:
\begin{lem} 
\label{lem:fp_covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Then it holds for some constant $C > 0$
\begin{equation}
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x)
\end{equation}
For the stationary distribution it holds
\begin{equation}
|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof}
Note that
\[
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq |\tilde{f}(x)| \mathsf{E}_x|\tilde{f}(X_s)| \leq \|\tilde{f}\|_{V^{\frac{1}{2}}}V^{\frac{1}{2}}(x)\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|
\]
By Hoelder inequality, 
\begin{eqnarray*}
\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)| & \leq & \left(\int\limits_{\rset^d} V(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2}\left(\int\limits_{\rset^d}|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2} \leq \\
& \leq & 2\rho^{s/2}V^{1/2}(x)
\end{eqnarray*}
yielding the first statement of lemma. The second statement can be obtained from the first one by integration with respect to $\pi$. 
\end{proof}

\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof} 
Note that
\[
\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] + \left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]
\]
Due to $V-$ergodicity, 
\[
\left|\left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{k+s}
\]
To bound the first term note that
\begin{eqnarray*}
\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| \leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)| \leq  \\
\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x)
\end{eqnarray*}
where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
\begin{lem}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Assume also that $n(1-\rho^{1/2}) > 1$. Then 
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})}) 
\end{equation}
\end{lem}
\begin{proof}
Note that 
\begin{eqnarray*}
\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}
\end{eqnarray*}
Now we bound first and second sum using lemma~\ref{lem:covariance}:
\[
S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)}
\]
\begin{eqnarray*}
S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[C^2V^2(x)\rho^{k+1}\frac{1}{1-\rho} + C^2V(x)\rho^{k+1/2}\frac{1}{1-\rho^{1/2}} + C\pi(V)\frac{1}{1-\rho^{1/2}}\|\tilde{f}\|_{V^{1/2}}\right] \leq \\
& \leq & \frac{2\rho^NC^2V^2(x)}{n^2(1-\rho)^2} + \frac{2\rho^NC^2V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})}
\end{eqnarray*}
Hence \eqref{eq:vanilla_var} follows.
\end{proof}

%Aforementioned lemmas allow us to bound the exponential moment of the additive functional of ULA. Namely, the following theorem holds
%
%\begin{thm}
%\label{thm:ula_const_size}
%Suppose that ULA kernel \ref{eq:kernel} satisfies assumptions {\bf (H1)} and {\bf (H2)}, and let $X_1,\ldots,X_n,\ldots$ be generated by ULA with constant step size $\gamma$. Let $X_0 = x$ be fixed. Then for any bounded function $g(x) : \mathbb{R}^d \rightarrow \mathbb{R}, |g(x)| \leq M$, it holds
%\begin{eqnarray*}
%\mathsf{E}\left[\exp\left(-\rho\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] \leq C_1\exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
%\end{eqnarray*}
%for some absolute constant $C_1$ not depending on $l,p$.
%\end{thm}
%
%\begin{proof} Conditions {\bf (H1)} and {\bf (H2)} imply that the Markov kernel $R_\gamma$ is $V-$ergodic (lemma \ref{lem:v_ergodicity}) and satisfies drift condition \ref{lem:drift}. Due to \cite[Theorem~3, Fact 3]{moulines:bound_dif:2019}, we obtain the following bound
%\begin{align*}
%\mathsf{E}_x\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)\right] \leq \exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
%\end{align*}
%\sergey{I will put precise constant later}
%with constant $\kappa = ...$. Note that
%\begin{align*}
%\mathsf{E}\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] =
%\end{align*}
%
%\begin{align*}
%=\mathsf{E}_x\left[\exp{\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)}\right]
%\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)}
%\end{align*}
%Using lemma~\ref{lem:v_ergodicity},
%\begin{align*}
